---
title: "cookbook.rmd"
author: "George Ofulu"
date: "10/24/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(magrittr)
library(dplyr)
library(stringr)
```

#### 1. Download the files into the rawData folder.

The following block of code checks if there's a `rawData` folder in the current working directory. If there's none, a folder is created and the [HAR data](https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip) from the [ UCI Machine Learning Repository](http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones) is downloaded. There are conditional statements to check if the data that's already downloaded is complete; if the data isn't completely downloaded, the file is downloaded and overwrites the previos `data.zip` file.

Note: At the time of the analysis, the dataset in the [ UCI Machine Learning Repository](http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones) was approximately 62.6 MB.
```{r, dataDonwload}
rawDataDownloadUrl <-
  "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"

# Condition 1: to check if the rawData folder exists
if (!dir.exists("./rawData")) {
  dir.create("./rawData")
}

# Condition 2: to check if the data has already been downloaded to avoid multiple downloads and also to check for an incomplete download.
if (!file.exists("./rawData/data.zip")) {
  download.file(rawDataDownloadUrl, destfile = "./rawData/data.zip")
  unzip("./rawData/data.zip", exdir = "./rawData")
} else if (file.size("./rawData/data.zip") < 62556944) {
  download.file(rawDataDownloadUrl, destfile = "./rawData/data.zip")
} else {
  print("Data Already exists in the target folder")
}
```


#### 2. Import useful files for the analysis into R

The R base package's `read.table` would be used to read in the following files:

- **The Test Data**: The file path is `"./rawData/UCI HAR Dataset/test/X_test.txt"`
- **The Test Data labels**: The file path is `"./rawData/UCI HAR Dataset/test/y_test.txt"`
- **Test Subject data**: The file path is `"./rawData/UCI HAR Dataset/test/subject_test.txt"`
- **The Train Data**: The file path is `"./rawData/UCI HAR Dataset/train/X_train.txt"`
- **The Train Data labels**: The file path is `"./rawData/UCI HAR Dataset/train/y_train.txt"`
- **Train Subject data**: The file path is `"./rawData/UCI HAR Dataset/train/subject_train.txt"`
- **Activity labels**: The file path is `"./rawData/UCI HAR Dataset/activity_labels.txt"`
- **Feature Names**: The file path is `"./rawData/UCI HAR Dataset/features.txt"`
```{r, dataImport}
testData <- read.table("./rawData/UCI HAR Dataset/test/X_test.txt", header = FALSE)
testDataLabel <- read.table("./rawData/UCI HAR Dataset/test/y_test.txt", header = FALSE)
testDataSubject <- read.table("./rawData/UCI HAR Dataset/test/subject_test.txt", header = FALSE)
trainData <- read.table("./rawData/UCI HAR Dataset/train/X_train.txt", header = FALSE)
trainDataLabel <- read.table("./rawData/UCI HAR Dataset/train/y_train.txt", header = FALSE)
trainDataSubject <- read.table("./rawData/UCI HAR Dataset/train/subject_train.txt", header = FALSE)
activityLabel <- read.table("./rawData/UCI HAR Dataset/activity_labels.txt", header = FALSE)
featureNames <- read.table("./rawData/UCI HAR Dataset/features.txt", header = FALSE)
featureNames <- as.vector(featureNames[[2]])
```

#### 3. Clean up Data

The train and test Data would be merged to their appropriate data labels and subject data to produce a dataframe with a two new columns to denote the activity type and subject. 

The dataframe is cross referenced with `featureNames` to produce a clean testData with column names that are meaningful.
```{r testDataCleanup}
testData <- data.frame(testDataSubject, testDataLabel, testData)
trainData <- data.frame(trainDataSubject, trainDataLabel, trainData)

names(testData) <- c("subjectID", "activityID", featureNames)
names(trainData) <- c("subjectID", "activityID", featureNames)
```

#### 4. Merge Train Data and Test Data
The `trainData` and `testData` would be merged to form one variable called `mergedHARData`.
```{r}
mergedHARData <- trainData %>% union(testData)
```

#### 5. Using Dplyr to select relevant columns in the Data and aggregating by Activity and Subject.
The `Select` verb from `dplyr` package to extracts only the measurements on the mean and standard deviation for each measurement.
```{r}
tidyHARData <- mergedHARData %>%
  select(subjectID, activityID, contains("mean()") | contains("std()")) %>%
  group_by(activityID, subjectID) %>%
  summarize(across(.fns = mean))
```
#### 6. Tidying up the dataset
The dataset would be tidied by tranforming the variable names to human readable variable names
```{r}
tidyHARDataNames <- names(tidyHARData)
tidyHARDataNames <- tidyHARDataNames %>%
  str_replace_all("-", "") %>% # remove hyphens
  str_replace_all("mean", "Mean") %>%
  str_replace_all("std", "StdDev") %>%
  str_replace_all("[(][)]", "") %>% # remove all parenthesis "()"
  str_replace_all("X", "Xaxis") %>%
  str_replace_all("Y", "Yaxis") %>%
  str_replace_all("Z", "Zaxis") %>%
  str_replace_all("tbody", "timeBody") %>%
  str_replace_all("fBody", "fourierTransformedBody")
names(tidyHARData) <- tidyHARDataNames
```

#### 7. Write the final data to a csv file

```{r}
write.table(tidyHARData, "tidyHARData.txt", row.names = FALSE)
```
